---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data
###First Load libraries, unzip and load data, then clean out the NA(s)
```{r}
library(ggplot2)
library(grid)
library(gridExtra)

filename <- "activity.zip"
datafilename <- "activity.csv"

activityData <- read.csv(unz(filename,datafilename),header=T)
activityData$steps <- as.numeric(activityData$steps)
activityData$interval <- as.numeric(activityData$interval)
activityData$date <- as.Date(activityData$date)

#get clean set of data
completeCaseActivity <- activityData[complete.cases(activityData),]
```


## What is mean total number of steps taken per day?
```{r}
dailyStepSum <- aggregate(completeCaseActivity$steps,list(date = completeCaseActivity$date),sum)


print(mean(dailyStepSum$x))

```

## What is the median number of steps taken per day?
```{r}
print(median(dailyStepSum$x))
```

## Histogram of daily activity
```{r histo_NAs_removed}
dailySumGraph=ggplot(data=dailyStepSum, aes(dailyStepSum$x)) + 
    geom_histogram( 
        col="black", 
        fill="green", 
        alpha = .4) + 
    labs(title="Histogram for sum of steps per day (NAs omitted") +
    labs(x="Number of Steps", y="Count") +
    ylim(c(0,10))+
    scale_y_continuous(breaks=c(0,2,4,6,8,10))+
    geom_vline(aes(xintercept=mean(dailyStepSum$x)),  
               color="red", linetype="dashed", size=1) +
    geom_text(aes(mean(dailyStepSum$x)+500,9,label = "mean"),angle=270)
print(dailySumGraph)
```

## What is the average daily activity pattern?

```{r averageDailyIntervalsGraph}
fiveMinMean <- aggregate (completeCaseActivity$steps,list(interval=completeCaseActivity$interval),mean)
names(fiveMinMean) <- c("interval","steps")

fiveMinMeanGraph = ggplot(fiveMinMean, aes(x = interval, y = steps)) + 
    geom_line(color = "black") + 
    labs(title = "Mean number of steps by Interval", x = "Interval", y = "Number of steps")
print(fiveMinMeanGraph)
```

###Max steps at:

```{r}
#find interval with the max number of steps
maxSteps <- which.max(fiveMinMean$steps)
print(fiveMinMean[maxSteps, ])
```

## Imputing missing values
In order to replace the NA values, we use the average of available data for that interval over the entire data set, rounded to the nearest whole number.  The round operation is done simply because the number we are representing is a whole number.  Specifically, a whole number representing the number of steps taken in that interval.

### Total number of missing values:
```{r}
## Get index of NAs
missingList <- which(is.na(activityData$steps))
print(length(missingList))
```

### Assign rounded average values for NA(s)
```{r}
completedData <- activityData
```

### Replace missing values 
```{r}
for(i in missingList) {
    completedData$steps[i] <- round(fiveMinMean[which(fiveMinMean$interval == completedData[i, "interval"]), "steps"])
    }
```

### Histogram of daily steps with NA(s) replaced
```{r histo_NAs_imputed}
#### Histogram with NAs FILLED
#get sum of daily steps for histogram
dailyStepSum <- aggregate(completedData$steps,list(date = completedData$date),sum)

dailySumFilledGraph=ggplot(data=dailyStepSum, aes(dailyStepSum$x)) + 
    geom_histogram( 
        col="black", 
        fill="green", 
        alpha = .4) + 
    labs(title="Histogram for sum of steps per day (NAs filled)") +
    labs(x="Number of Steps", y="Count") +
    ylim(c(0,10))+
    scale_y_continuous(breaks=c(0,2,4,6,8,10))+
    geom_vline(aes(xintercept=mean(dailyStepSum$x)),  
               color="red", linetype="dashed", size=1) +
    geom_text(aes(mean(dailyStepSum$x)+500,9,label = "mean"),angle=270)

print(dailySumFilledGraph)
```

####Mean of daily activity with NA(s) replaced
```{r}
print(mean(dailyStepSum$x))
```

####Median of daily activity with NA(s) replaced
```{r}
print(median(dailyStepSum$x))
```


## Are there differences in activity patterns between weekdays and weekends?
### Primarily the difference is in a later start time for activities, and less well defined activity spikes at specific times

```{r comparisonGraph}
completedData[as.POSIXlt(completedData$date)$wday %in% c(0,6),"dayType"]="weekend"
completedData[as.POSIXlt(completedData$date)$wday %in% c(1,2,3,4,5),"dayType"]="weekday"
completedData$dayType <- as.factor(completedData$dayType)


completedDataByInterval <- aggregate(completedData$steps,
                                     by = list(completedData$interval, completedData$dayType),
                                     FUN = mean)

## Name columns
names(completedDataByInterval) <- c("interval","dayType","steps")

g<-ggplot(completedDataByInterval,aes(interval,steps))
p<-g+geom_line()+geom_point()+facet_grid(dayType~.)
p<-p+labs(title="Average steps in interval for weekend vs weekday")
p<-p+scale_x_continuous(breaks=c(400,800,1200,1600,2000,2400))
print(p)
```
